Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2022-06-20 analysis/paper_draft2.Rmd
revised text; figure and equation cross-references

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  #thanks: The authors gratefully acknowledge ...
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
    #fig_width: 5
    #fig_height: 5
  #output:
#  pdf_document:
#    latex_engine: xelatex
#    number_sections: true
    #fig_width: 4
    #fig_height: 4
header-includes: 
  - \usepackage{amsmath}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, gender, or membership in certain other cognizable classes. Here, we propose a Bayesian approach for detecting such illegal bias. We show how to incorporate historical data on an attorney's use of peremptory strikes in past cases to estimate how likely that attorney has struck prospective jurors in a current trial because of their cognizable-class membership. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. We show how well our model performs on simulated data. Finally, we extend this approach with a prototype software application to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with historical strike data from a convenience sample of cases from one court.

@@

In the U.S., individuals selected for jury-service appear in court as scheduled and are questioned by the parties' attorneys and the trial judge. During this process, a prospective juror, if not excused by the trial judge for cause, may still be dismissed if a party's attorney uses one of their limited number of peremptory challenges against them. By asserting a peremptory challenge, a party can declare a prospective juror ineligible ("strikes" that juror) for a seat on the jury without the burden of explaining why.

Since _Batson v. Kentucky_ [-@batson1986], a party violates the Equal Protection Clause of the US Constitution by using peremptory challenges if motivated by the prospective jurors' race, ethnicity, gender, or certain other cognizable classes. The party bringing a _Batson_ challenge bears the burden of proving such illegal bias is more likely than not to be true [@LaFave2020, $\S$ 22.3(d)]. For similar challenges under State law, a few States require only that an "objective observer" or an "objectively reasonable person" would find that race, ethnicity or another cognizable-class was a "factor" in that party's use of strikes [Wash. General Rule. 37(e); Calif. Code Civil Procedure $\S$ 231.7(d)(1)]. Peremptory-challenge procedure, and thus the task of proving such illegal bias, varies not only by State, but also by court, including the number of peremptory challenges assigned to each side and the order in which each party uses those strikes [@nscs-jury2018; @Williams2017].

In any case, evidence of such illegal bias may include data on a party's use of peremptory challenges in past cases [@flowers2019, 2243; Wash. General Rule 37(g)(v); Calif. Code Civil Procedure $\S$ 231.7(d)(3)(G)]. Prior studies have collected historical strike data in past cases and reported the observed difference in strike rates by race or gender. Typically, they test for the probability of observing a non-zero difference in strike rates by the race or gender of the struck prospective jurors, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2021.

@@

To evaluate the proposed method, we conducted a simulation study using Stan `r rstan::stan_version()`, a Hamiltonian Monte Carlo engine for Bayesian inference by way of `r R.version$version.string` and RStan [@rstan]. <!--To measure convergence, we used trace plots, the Gelman-Rubin convergence diagnostic and Geweke’s diagnostics [@Gelman1992;@Gelman1998;@Geweke1992].--> These simulations primarily show how much our ability to detect bias was affected the size of the historical data and by how much the power prior ($\alpha$) discounted the weight of the historical data. As introduced above, $b > 0$ represents a prosecutor's bias against a prospective juror within the cognizable class, while $b = 0$ denotes no bias. For simplicity, we assume the defense attorney has no bias.

Because we use the power prior, we can control how much we account for the historical strike data by modifying the discounting parameter $\alpha$. If $\alpha=1$, the historical strike data is equally weighted with the data on strikes in the current trial. If $\alpha<1$, the historical data are discounted and weighted proportionally less than the current trial data. We assign $\alpha<1$ to evaluate how sensitive the posterior of the bias parameter is to the historical data on strikes. Accordingly, in the simulation study, we evaluated different values of $\alpha$: $\alpha=\{0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1\}$ to show how that parameter affected bias detection .

@@

To generate data, we generated strike data for a current trial using seven different values of the bias parameter $b_{curr}=\{0,0.5,1,1.5,2, 2.5,3\}$. We selected these values to evaluate the strength of the power prior as a function of the compatibility between current and historical strike data $\mid b_{curr} - b_{hist} \mid$. Given perfect compatibility, $\mid b_{curr} - b_{hist} \mid = 0$. To allow for imperfect compatibility, we let $b_{hist}$ to have the same selected bias values as $b_{curr}$, i.e., $b_{hist}=\{0,0.5,1,1.5,2, 2.5,3\}$. We did this to simulate noise in the historical strike data that can arise due to not-at-random missingness or because a confounder (e.g. charge severity) affects $b_curr$ but not of all the past trials in the historical strike data.

For our simulation, we define bias detection as the proportion of times we identify bias when bias is present in the current trial, i.e., $b_{curr} > 0$ , based on whether credible intervals of the bias parameter exclude zero. Since we use a weakly-informative prior that pulls the estimate of $b_{curr}$ towards zero, we focus on how accurately our model can detect that bias, _not_ on how accurately the model can recover the true value of $b_{curr}$ (i.e., a traditional coverage rate).  For instance, a 90\% bias detection rate means that in 900 of 1000 model fits, a given credible interval does not contain zero if the $b_{curr}$ is not zero. Put another way, if we set $b_{curr}>0$, we calculate the proportion of the 90\% credible intervals that lie to the right of zero (lower bound is positive) among the 1000 model fits. For each scenario, we generated 1000 data sets and fit the model on those data sets. For each replicate, we generate a MCMC sample of 10,000 iterations with a burn-in period of 1000 iterations. For each replicate, we calculate the posterior mean and $80\%$, $90\%$, and $95\%$ highest posterior density (HPD) intervals for $b_{curr}$.

Additionally, we present sensitivity analysis for the number of available strikes with 6 strikes, 10 strikes, and 15 strikes for each attorney. Thus, in total, we considered 4851 different scenarios, i.e. the combination of 7 different bias parameters of current data and of historical data, 3 different amounts of historical data, 11 different values for the power prior weight parameter, and for 3 combinations of the total number of strikes.

@@

Figure \@ref(fig:figbd9515) depicts the bias detection rate with an 95\% credible interval for the 15 strikes scenario. When $b_{curr}$ is high (close to 3) the coverage rate is high, very close to 1. However, when $b_{curr}$ is lower (close to zero), the bias detection rate drops below 0.8. When there is no bias in the current or historical trials ($b_{curr} = b_{hist} = 0$), our bias detection rate for 95% credible interval is 0.05 or less. As we place more weight on the historical strike data (as $\alpha$ increases), and the historical strike data is consistent with current information (high compatibility), the bias detection rate is close to 1, especially when bias is high (upper right-hand corner of plots). Conversely, as $\alpha$ decreases, bias detection is only high when there is high bias in the current trial ($b_{curr} \geq 1.5$). Increasing the number of historical trials leads to an increase in bias detection especially when current and historical trials are compatible.

@@

As we decrease the number of strikes each attorney has, from fifteen to ten to six (Figures \@ref(fig:figbd9510) and \@ref(fig:figbd9506) respectively), our ability to detect bias decreases. Bias detection is only high when $\alpha$ is high (close to 1), current trial bias is high, there is high compatibility between current and historical trials, and the number of historical trials is large.

@@

Bias detection will also be sensitive to the choice of the standard deviation of the prior. The results above depend in part on our choice a conservative initial prior for the bias parameter ($b \sim N(0,2)$). Given little historical strike data for an attorney and a low number of strikes per trial, this initial prior will dominate. As a result, the model we use will only detect severe bias. This sensitivity, however, reduces after including the historical data, especially when $\alpha$ is close to 1.

@@

Figure \@ref(fig:figsdprior) depicts how different values of the initial prior’s standard deviation ($\sigma = \{1,2,100\}$) affects the bias detection rate for three $\alpha$ values ($\alpha = \{0.1,0.5,1\}$). If we set initial prior's standard deviation to 1 (half the value used in our simulations), that matters most when $\alpha$ is 0.1, i.e., borrowing less information from the historical data. As $\alpha$ increases, the initial prior's standard deviation matters less. Moreover, increasing the initial prior's standard deviation -- thus making the initial prior less informative -- has little impact on bias detection rates, regardless of choice of $\alpha$.

@@

## Simultaneous Strikes

@@

Moreover, suppose an attorney of interest exercised strikes in two courts, each with different strike procedures. In some cases, one court's strike procedure can be modeled as a special case of a more general model that can be applied to another court's procedure (e.g., the simultaneous strike example above in section \@ref(Simultaneous-Strikes 
)). In other cases, however, one cannot apply the same model to courts with different jury selection practices. For example, the model we used here does not apply to strike data generated by a court's strike procedure in which prospective jurors appear before the attorneys one at a time to be either struck or seated on the jury.

Furthermore, the bias detection rate depends on the degree of incompatibility between the bias parameters for the current trial and for the past (historical) trials ($\mid b_{curr} - b_{hist} \mid$). These bias parameters may be incompatible ($b_{curr} \neq b_{hist}$) for at least two reasons.

@@

In practice, the anticipated degree of incompatibility should influence which value of $\alpha$ to select. For example, given perfect compatibility ($\mid b_{curr} - b_{hist} \mid = 0$), we should let $\alpha = 1$ (equal weight between current and historical data). This will maximize bias detection even when the number of strikes per trial is small. This effect of increased bias detection can be seen in the simulation results depicted in Figures \@ref(fig:figbd9515), \@ref(fig:figbd9510), and \@ref(fig:figbd9506) along the positive diagonals of the individual blocks.

@@

In this paper, we extended a Bayesian approach to estimate attorney strike bias in the use of peremptory challenges by incorporating historical data on that attorney's use of strikes in past cases. In so doing, we used the power prior to adjust the weight of such historical information. Our simulations showed that how well our approach detects attorney bias depending on the number of past trials in which that attorney used strikes; the number of strikes per trial; the degree of compatibility, i.e., the distance between the bias parameters for the current trial and for the past trials. Finally, we discussed a prototype software application with which attorneys could use our approach in real time during jury selection to detect attorney strike bias.

>>>>

---
title: "Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data"
date: "`r Sys.Date()`"
blinded: 0
authors:
- name: "[Sachin S. Pandya](https://orcid.org/0000-0001-7387-1307)"
  #thanks: The authors gratefully acknowledge ...
  affiliation: University of Connecticut
  email: sachin.pandya@uconn.edu
- name: "Xiaomeng Li"
  affiliation: University of Connecticut
- name: "[Eric Barón](https://orcid.org/0000-0003-1934-1932)"
  affiliation: University of Connecticut
- name: "[Timothy E. Moore](https://orcid.org/0000-0002-9576-0517)"
  affiliation: University of Connecticut
output:
  bookdown::pdf_book: 
      base_format: rticles::asa_article
      citation_package: "default"
      keep_tex: TRUE
header-includes: 
  - \usepackage{amsmath}
#fontsize: 12pt
bibliography: references.bib
biblio-style: agsm.bst
nocite: |
  @WashR37, @calccp2021
keywords: "Batson challenge, peremptory strikes, power prior, Bayesian"
abstract: |
  US law bars using peremptory strikes during jury selection because of prospective juror race, ethnicity, gender, or membership in certain other cognizable classes. Here, we propose a Bayesian approach for detecting such illegal bias. We show how to incorporate historical data on an attorney's use of peremptory strikes in past cases to estimate how likely that attorney has struck prospective jurors in a current trial because of their cognizable-class membership. In so doing, we use the power prior to adjust the weight of such historical information in the analysis. We show how well our model performs on simulated data. Finally, we extend this approach with a prototype software application to detect bias in peremptory-challenge in real time during jury-selection. We illustrate this application's use with historical strike data from a convenience sample of cases from one court.

@@

In the United States, individuals selected for jury-service appear in court as scheduled and are questioned by the parties' attorneys and the trial judge. During this process, a prospective juror, if not excused by the trial judge for cause, may still be dismissed if a party's attorney uses one of their limited number of peremptory challenges against them. By asserting a peremptory challenge, a party can declare a prospective juror ineligible ("strikes" that juror) for a seat on the jury without the burden of explaining why.

Since _Batson v. Kentucky_ [-@batson1986], a party violates the Equal Protection Clause of the US Constitution by using peremptory challenges if motivated by the prospective jurors' race, ethnicity, gender, or certain other cognizable classes. The party bringing a _Batson_ challenge bears the burden of proving such illegal bias is more likely than not to be true [@LaFave2022, $\S$ 22.3(d)]. For similar challenges under State law, a few States require only that an "objective observer" or an "objectively reasonable person" would find that race, ethnicity or another cognizable-class was a "factor" in that party's use of strikes [Wash. General Rule. 37(e); Calif. Code of Civil Procedure $\S$ 231.7(d)(1)]. Peremptory-challenge procedure, and thus the task of proving such illegal bias, varies not only by State, but also by court, including the number of peremptory challenges assigned to each side and the order in which each party uses those strikes [@nscs-jury2018; @Williams2017].

In any case, evidence of such illegal bias may include data on a party's use of peremptory challenges in past cases [@flowers2019, 2243; Wash. General Rule 37(g)(v); Calif. Code Civil Procedure $\S$ 231.7(d)(3)(G)]. Prior studies have collected historical strike data in past cases and reported the observed difference in strike rates by race or gender. Typically, they test for the probability of observing a non-zero difference in strike rates by the race or gender of the struck prospective jurors, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014, 289-297; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2021.

@@

To evaluate the proposed method, we conducted a simulation study using Stan `r rstan::stan_version()`, a Hamiltonian Monte Carlo engine for Bayesian inference by way of `r R.version$version.string` and RStan [@rstan]. <!--To measure convergence, we used trace plots, the Gelman-Rubin convergence diagnostic and Geweke’s diagnostics [@Gelman1992;@Gelman1998;@Geweke1992].--> These simulations primarily show how much our ability to detect bias was affected the size of the historical data and by how much the power prior discounted the weight of the historical data. As introduced above, $b > 0$ represents a prosecutor's bias against a prospective juror within the cognizable class, while $b = 0$ denotes no bias. For simplicity, we assume the defense attorney has no bias.

Because we use the power prior, we can control how much we account for the historical strike data by modifying the discounting parameter $\alpha$ (denoted as $\alpha_0$ in Equations \@ref(eq:powerprior) and \@ref(eq:model4)). If $\alpha=1$, the historical strike data is equally weighted with the data on strikes in the current trial. If $\alpha<1$, the historical data are discounted and weighted proportionally less than the current trial data. We assign $\alpha<1$ to evaluate how sensitive the posterior of the bias parameter is to the historical data on strikes. Accordingly, in the simulation study, we evaluated different values of $\alpha$: $\alpha=\{0,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1\}$ to show how that parameter affected bias detection.

@@

To generate data, we generated strike data for a current trial using seven different values of the bias parameter $b_{curr}=\{0,0.5,1,1.5,2, 2.5,3\}$. To evaluate the strength of the power prior as a function of the degree of incompatibility between current and historical strike data ($\lvert b_{curr} - b_{hist} \rvert > 0$), we assigned $b_{hist}$ the same bias values as $b_{curr}$, i.e., $b_{hist}=\{0,0.5,1,1.5,2, 2.5,3\}$. We did this to simulate noise in the historical strike data that can arise due to not-at-random missingness or because a confounder (e.g. charge severity) affects $b_curr$ but not of all the past trials in the historical strike data.

For our simulation, we define bias detection as the proportion of times we identify bias when bias is present in the current trial, i.e., $b_{curr} > 0$ , based on whether credible intervals of the bias parameter exclude zero. Since we use a weakly-informative prior that pulls the estimate of $b_{curr}$ towards zero, we focus on how accurately our model can detect that bias, _not_ on how accurately the model can recover the true value of $b_{curr}$ (i.e., a traditional coverage rate).  For instance, a 90\% bias detection rate means that in 900 of 1000 model fits, a given credible interval does not contain zero if the $b_{curr}$ is not zero. Put another way, if we set $b_{curr}>0$, we calculate the proportion of the 90\% credible intervals that lie to the right of zero (lower bound is positive) among the 1000 model fits. For each scenario, we generated 1000 data sets and fit the model on those data sets. For each replicate, we generate a MCMC sample of 10,000 iterations with a burn-in period of 1000 iterations. For each replicate, we calculate the posterior mean and $80\%$, $90\%$, and $95\%$ highest posterior density intervals for $b_{curr}$.

Additionally, we present sensitivity analysis for the number of available strikes per trial (6 strikes, 10 strikes, and 15 strikes for each attorney). Thus, in total, we considered 4851 different scenarios, i.e. the combination of 7 different bias parameters of current data and of historical data, 3 different amounts of historical data, 11 different values for the power prior weight parameter, and for 3 combinations of the total number of strikes.

@@

Figure \@ref(fig:figbd9515) depicts the bias detection rate with an 95\% credible interval for the 15 strikes scenario. When $b_{curr}$ is high (close to 3) the bias detection rate is high (very close to 1). However, when $b_{curr}$ is lower (close to zero), the bias detection rate drops below 0.8. When there is no bias in the current or historical trials ($b_{curr} = b_{hist} = 0$), the bias detection rate for the 95% credible interval is 0.05 or less. As we place more weight on the historical strike data (as $\alpha$ increases), and the historical strike data is consistent with current information (high compatibility), the bias detection rate is close to 1, especially when bias is high (upper right-hand corner of plots). Conversely, as $\alpha$ decreases, bias detection is only high when there is high bias in the current trial ($b_{curr} \geq 1.5$). Increasing the number of historical trials leads to an increase in bias detection especially when current and historical trials are compatible.

@@

As we decrease the number of strikes each attorney has, from fifteen to ten to six (Figures \@ref(fig:figbd9510) and \@ref(fig:figbd956) respectively), bias detection decreases. Bias detection is only high when $\alpha$ is high (close to 1); current trial bias is high; there is high compatibility between current and historical trials; and the number of historical trials is large.

@@

Bias detection will also be sensitive to the choice of the standard deviation of the prior. The results above depend in part on our choice of a conservative initial prior for the bias parameter ($b \sim N(0,2)$). Given little historical strike data for an attorney and a low number of strikes per trial, this initial prior will dominate. As a result, the model we use will only detect severe bias. This sensitivity, however, reduces after including the historical data, especially when $\alpha$ is close to 1.

@@

Figure \@ref(fig:figsdprior) depicts how different values of the initial prior’s standard deviation ($\sigma = \{1,2,100\}$) affects the bias detection rate for three $\alpha$ values ($\alpha = \{0.1,0.5,1\}$). In our simulations, we set $\sigma = 2$ for our initial prior. If we instead select $\sigma = 1$, that matters most when $\alpha = 0.1$, i.e., when we borrow less information from the historical data. As $\alpha$ increases, the initial prior's standard deviation matters less. Moreover, when we increased the initial prior's standard deviation to make it much less informative ($\sigma = 100$), that had little impact on bias detection rates, regardless of the choice of $\alpha$.

@@

## Simultaneous Strikes {#simul-strikes}

@@

Moreover, suppose an attorney of interest exercised strikes in two courts, each with different strike procedures. In some cases, one court's strike procedure can be modeled as a special case of a more general model that can be applied to another court's procedure (e.g., the simultaneous strike example in section \@ref(simul-strikes) ). In other cases, however, one cannot apply the same model to courts with different jury selection practices. For example, the model we used here does not apply to strike data generated by a court's strike procedure in which prospective jurors appear before the attorneys one at a time to be either struck or seated on the jury.

Furthermore, the bias detection rate depends on the degree of incompatibility between the bias parameters for the current trial and for the past (historical) trials ($\lvert b_{curr} - b_{hist} \rvert$). These bias parameters may be incompatible ($b_{curr} \neq b_{hist}$) for at least two reasons.

@@

In practice, the anticipated degree of incompatibility should influence which value of $\alpha$ to select. For example, given perfect compatibility ($\lvert b_{curr} - b_{hist} \rvert = 0$), we should let $\alpha = 1$ (equal weight between current and historical data). This will maximize bias detection even when the number of strikes per trial is small. This effect of increased bias detection can be seen in the simulation results depicted in Figures \@ref(fig:figbd9515), \@ref(fig:figbd9510), and \@ref(fig:figbd956) along the positive diagonals of the individual blocks.

@@

In this paper, we extended a Bayesian approach to estimate attorney strike bias in the use of peremptory challenges by incorporating historical data on that attorney's use of strikes in past cases. In so doing, we used the power prior to adjust the weight of such historical information. Our simulations showed that how well our approach detects attorney bias depends on the number of past trials in which that attorney used strikes; the number of strikes per trial; and the degree of incompatibility, i.e., the distance between the bias parameters for the current trial and for the past trials. Finally, we discussed a prototype software application with which attorneys could use our approach in real time during jury selection to detect attorney strike bias.
